---
title: "Data 101 Final Paper"
output: github_document
---
```{r}
install.packages("tidyverse")
library(tidyverse)
```
```{r}
install.packages("psych")
library(psych)
dataset <- read.csv("Data 101 Final Project Data Cleaned.csv")
```
# Section 1 - Introduction:

Many people are becoming more extreme with their political beliefs in recent years, moving away from moderate positions and towards either the left or right of the spectrum. This has raised concern about hardening of policy positions. Ideally, people would change their policy positions when presented with new situations, but extremism might prevent their position from changing. We were motivated to research this topic because of its impact on policy and cooperation in American politics.

Our principal research question is: Does a person’s extremism, defined as distance from a politically neutral position, affect the level of absolutism, defined as a person’s propensity to maintain their position in extreme scenarios? Is this relationship true for only social (abortion, immigration, gun control) or only economic (tariffs, taxes) issues?

We used a dataset of survey responses on Prolific measuring extremism and absolutism. The questions relate to various economic and social issues. 

# Section 2 - Data:

The data was collected via Prolific, a national survey platform. There were 41 survey respondents (3 responses were removed). The original dataset included 45 variables, plus 20 variables created for this project. The final analysis includes 38 observations of 12 variables. 

Participants answered questions on gun control, abortion, immigration, tariffs, and taxes. Each  included an initial question to determine participants’ opinions (scale of 1-6). Participants were given a set of three statements, such as “[preferred policy] would always be right, even in a county where no one thought it was right” based on their original opinion, and could answer either ‘I agree’ or ‘I disagree’. Each participants’ score on all three questions were summed so that a 0 meant they had disagreed with everything, and a 3 meant they had agreed with everything. 

This data’s main limitation is the low number of participants, making it difficult to make proper inferences.

# Section 3 - Analytic Framework

Our explanatory variable is extremism. The extremism variable was created by recoding the initial question gauging opinion on each issue. A 1 or 6 was recoded to a 3, a 2 or 5 to a 2, and a 3 or 4 to a 1, so that an opinion further away from the middle of the scale (more extreme) is higher. This created an extremism score for each issue, which were averaged together across all issues as well as social and economic issues. This variable is continuous on a scale from 1 - 3.

Our response variable is absolutism. Each participant had a ‘moral absolutism’ score for each issue (described above), which were then averaged together across all issues to create an aggregate absolutism score. The scores on the social issues (gun control, abortion, immigration) and the economic issues (tariffs, taxes) were also averaged together separately. This variable is also continuous on a scale of 0 - 3.

We hypothesized that holding extreme political positions is positively correlated with a high degree of absolutism. We expect that this relationship will be weaker in cases of economic issues, as economic issues tend to be more complicated and less polarizing. 

We built regression models to predict the relationship between extremism and absolutism. First, we tested the relationship between extremism (explanatory variable) and absolutism (response variable) at the aggregate level to check for an overall effect with a regression analysis.  Then we tested this relationship at an economic level and a social level.

# Section 4 - Results:

The summary stats for extremism and absolutism at the aggregate, economic, and social levels are provided below, as well as bar graphs for each variable across all levels. 

```{r}
dataset1 <- dataset |>
  mutate(SocialAbs = (GunControlAbs + AbortionAbs + ImmigrationAbs)/3) |>
  mutate(EconomicAbs = (TaxesAbs + TariffsAbs)/2) |>
  mutate(SocialExtr = (GunControlExtr + AbortionExtr + ImmigrationExtr)/3) |>
  mutate(EconomicExtr = (TaxesExtr + TariffsExtr)/2)
summarydata <- dataset1 |>
  select(AbsolutismAvg:EconomicExtr, ImmigrationAbsNEW, GunControlAbs:TariffsExtr)
summarydataframe <- na.omit(summarydata)
descriptivestats <- describe(summarydataframe)
print(descriptivestats)
```
```{r}
meansdataframe <- summarydataframe |>
  summarise_all(mean, na.rm = TRUE)
meansdatalongvis1 <- meansdataframe |>
  pivot_longer(cols = GunControlAbs:TariffsAbs,
               names_to = "Issues",
               values_to = "Means")
meansdatalongvis1 |>
  ggplot(
    aes(Issues, Means)
  )+
  geom_col()
```
```{r}
meansdatalongvis2 <- meansdataframe |>
  pivot_longer(cols = GunControlExtr:TariffsExtr,
               names_to = "Issues",
               values_to = "Means")
meansdatalongvis2 |>
  ggplot(
    aes(Issues, Means)
  )+
  geom_col()
```
```{r}
meansdatalongvis3 <- meansdataframe |>
  pivot_longer(cols = c(AbsolutismAvg, SocialAbs, EconomicAbs),
               names_to = "Aggregates",
               values_to = "Means")
meansdatalongvis3 |>
  ggplot(aes(Aggregates, Means))+
  geom_col()
```
```{r}
meansdatalongvis4 <- meansdataframe |>
  pivot_longer(cols = c(ExtremismAvg, SocialExtr, EconomicExtr),
               names_to = "Aggregates",
               values_to = "Means")
meansdatalongvis4 |>
  ggplot(aes(Aggregates, Means))+
  geom_col()
```

Additionally, a scatterplot with a line of best fit can be seen below for each level.

```{r}
summarydataframe |>
  ggplot(
    aes(SocialAbs, SocialExtr)
  )+
  geom_point()+
  geom_smooth(method = "lm")
```
```{r}
summarydataframe |>
  ggplot(
    aes(EconomicAbs, EconomicExtr)
  )+
  geom_point()+
  geom_smooth(method = "lm")
```
```{r}
summarydataframe |>
  ggplot(
    aes(AbsolutismAvg, ExtremismAvg)
  )+
  geom_point()+
  geom_smooth(method = "lm")
```

These visualizations seem to suggest a positive trend between extremism and absolutism. At the issue level, lower extremism scores are associated with lower absolutism scores, although there was less variation in extremism scores across the five issues than absolutism. The same can be said on the aggregate level versus the social and economic averages.

The scatterplots all display positive trends as well, though the line of best fit seems the most different from zero for the Economic issues. This appears to counter our original hypothesis that Social issues would display a more pronounced relationship between extremism and absolutism.

Based on the residual plots (below), we have met the conditions for a regression analysis. There does not seem to be any strong patterns or clusters in the residuals. However, regression results should still be treated cautiously due to the small number of observations.

```{r}
aggregatemodel <- lm(AbsolutismAvg ~ ExtremismAvg, summarydataframe)
socialmodel <- lm(SocialAbs ~ SocialExtr, summarydataframe)
economicmodel <- lm(EconomicAbs ~ EconomicExtr, summarydataframe)

summarydataframe$aggregateresid <- residuals(aggregatemodel)
summarydataframe$socialresid <- residuals(socialmodel) 
summarydataframe$economicresid <- residuals(economicmodel)
summarydataframe$aggregatefitted <- fitted(aggregatemodel) 
summarydataframe$socialfitted <- fitted(socialmodel) 
summarydataframe$economicfitted <- fitted(economicmodel)
```
```{r}
summarydataframe |>
  ggplot(aes(aggregatefitted, aggregateresid))+
  geom_point()+
  geom_hline(yintercept = 0)
```
```{r}
summarydataframe |>
  ggplot(aes(socialfitted, socialresid))+
  geom_point()+
  geom_hline(yintercept = 0)
```
```{r}
summarydataframe |>
  ggplot(aes(economicfitted, economicresid))+
  geom_point()+
  geom_hline(yintercept = 0)
```
The regression results below support our primary hypothesis, that extremism is a predictor of absolutism (p = .028). The regression line is: 
Absolutism = 1.048 + .378*Extremism
Although the slope is significantly different from zero, the variation in extremism only accounts for 12.65% of the variation in absolutism, though this may be due to the small amount of data.

```{r}
summary(aggregatemodel)
```
For the social issues, extremism does not have a significantly significant effect on absolutism. 

```{r}
summary(socialmodel)
```
For economic issues, extremism was a significant predictor of absolutism (p=.04). The regression equation is:
Absolutism = .916 + .486*Extremism
This indicates a positive relationship. However, the variation in extremism accounts for only 11.21% of the variation in absolutism in this model as well, though this may again be due to the limitations introduced by a smaller number of observations.

```{r}
summary(economicmodel)
```
Overall, extremism was associated with absolutism across all issues, but was not for social issues. This does not support our hypothesis that social issues would display a stronger relationship between the two variables than economic issues. Although the aggregate and economic models were significant, the R^2 values were not very high, and our ability to make definite inferences is limited by the small dataset.

# Section 5 - Implications

In the face of more extreme situations and a more polarized climate, this data illustrates how US citizens are more likely to keep to their current political view even in the face of extreme political scenarios that could change their mind. This trend is more likely to hold true for economic issues (tariffs and taxes) than social issues (abortion, immigration, and gun control). While it is useful to know the importance of extremism on absolutism with the current issues society is facing, it is difficult to make a generalized conclusion based on this data due to the small sample size. More survey responses need to be collected and analyzed to better understand this relationship and how it applies to the general US public.
